---
title: "Hands-on Exercise 4"
author: "Jin Zhang"
date: "8 May 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: false
---

# 8 Funnel Plots for Fair Comparisons

## 8.1 Overview

Funnel plots are designed for conducting **unbiased comparison** between outlets, stores or business entities. In this section we will learn to:

- Plot funnel plots using **FunnelPlotR**
- Plot static funnel plots using **ggplot2**
- Plot interactive funnel plots using **plotly** and **ggplot2**

## 8.2 Installing and Loading R Packages

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

## 8.3 Importing Data

The dataset used is `COVID-19_DKI_Jakarta.csv`, containing cumulative COVID-19 cases and deaths by sub-district as at 31 July 2021.

```{r}
covid19 <- read_csv("../../data/COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

```{r}
kable(head(covid19), format = "html")
```

## 8.4 FunnelPlotR Methods

### 8.4.1 Basic Plot

```{r}
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

### 8.4.2 Makeover 1 - Change data_type and set axis ranges

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05)
)
```

### 8.4.3 Makeover 2 - Add titles and remove outlier labels

```{r}
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label = "Cumulative COVID-19 Positive Cases",
  y_label = "Cumulative Fatality Rate"
)
```

## 8.5 Funnel Plot with ggplot2

### 8.5.1 Computing Derived Fields

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate * (1 - rate)) / Positive)) %>%
  filter(rate > 0)

fit.mean <- weighted.mean(df$rate, 1 / df$rate.se^2)
```

### 8.5.2 Calculating 95% and 99.9% Confidence Interval Limits

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean * (1 - fit.mean)) / number.seq)

dfCI <- data.frame(
  number.seq, number.ll95, number.ul95,
  number.ll999, number.ul999, fit.mean
)
```

### 8.5.3 Static Funnel Plot

```{r}
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label = `Sub-district`),
             alpha = 0.4,
             size = 0.9) +
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ll95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ul95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ll999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_line(data = dfCI,
            aes(x = number.seq, y = number.ul999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_hline(data = dfCI,
             aes(yintercept = fit.mean),
             linewidth = 0.4,
             colour = "grey40") +
  coord_cartesian(ylim = c(0, 0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") +
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") +
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") +
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size = 12),
        legend.position = c(0.91, 0.85),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.background = element_rect(colour = "white"),
        legend.key = element_blank())

p
```

### 8.5.4 Interactive Funnel Plot with plotly

```{r}
ggplotly(p, tooltip = c("label", "x", "y"))
```

---

# 9 Visualising Distribution

## 9.1 Learning Outcome

In this section, we explore two relatively new statistical graphic methods:

- **Ridgeline plots** using `ggridges`
- **Raincloud plots** using `ggdist`

## 9.2 Getting Started

### 9.2.1 Installing and Loading Packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

### 9.2.2 Importing Data

```{r}
exam <- read_csv("../../data/Exam_data.csv")
```

## 9.3 Ridgeline Plot

### 9.3.1 Basic Ridgeline Plot using `geom_density_ridges()`

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(name = "English grades",
                     expand = c(0, 0)) +
  scale_y_discrete(name = NULL,
                   expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

### 9.3.2 Varying Fill Colours Along the x-axis

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(name = "English grades",
                     expand = c(0, 0)) +
  scale_y_discrete(name = NULL,
                   expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

### 9.3.3 Mapping Probabilities onto Colour using `stat_density_ridges()`

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5 - stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient",
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

### 9.3.4 Ridgeline Plot with Quantile Lines

```{r}
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile)))) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

## 9.4 Raincloud Plot

### 9.4.1 Half Eye Graph

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

### 9.4.2 Adding Boxplot

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = 0.20,
               outlier.shape = NA)
```

### 9.4.3 Adding Dot Plots with `stat_dots()`

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = 0.20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = 0.5,
            dotsize = 2)
```

### 9.4.4 Finishing Touch

```{r}
ggplot(exam,
       aes(x = RACE,
           y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = 0.20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = 0.5,
            dotsize = 1.5) +
  coord_flip() +
  theme_economist()
```

---

# 10 Visual Statistical Analysis

## 10.1 Learning Outcome

In this section we use **ggstatsplot** to create visual graphics with rich statistical information embedded directly in the plot.

## 10.2 Getting Started

### 10.2.1 Installing and Loading Packages

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

### 10.2.2 Importing Data

```{r}
exam <- read_csv("../../data/Exam_data.csv")
```

## 10.3 One-Sample Test: `gghistostats()`

```{r}
set.seed(1234)

ggstatsplot::gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

## 10.4 Two-Sample Mean Test: `ggbetweenstats()`

```{r}
ggstatsplot::ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

## 10.5 One-way ANOVA Test: `ggbetweenstats()`

```{r}
ggstatsplot::ggbetweenstats(
  data = exam,
  x = RACE,
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

## 10.6 Significant Test of Correlation: `ggscatterstats()`

```{r}
ggstatsplot::ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE
)
```

## 10.7 Significant Test of Association: `ggbarstats()`

```{r}
exam1 <- exam %>%
  mutate(MATHS_bins =
           cut(MATHS,
               breaks = c(0, 60, 75, 85, 100)))

ggstatsplot::ggbarstats(exam1,
                        x = MATHS_bins,
                        y = GENDER)
```

---

# 11 Visualising Uncertainty

## 11.1 Learning Outcome

In this section we will learn to:

- Plot error bars using **ggplot2**
- Plot interactive error bars combining **ggplot2**, **plotly** and **DT**
- Create advanced uncertainty visualisations using **ggdist**
- Create **Hypothetical Outcome Plots (HOPs)** using **ungeviz**

## 11.2 Getting Started

### 11.2.1 Installing and Loading Packages

```{r}
pacman::p_load(plotly, crosstalk, DT,
               ggdist, ggridges, colorspace,
               gganimate, tidyverse)
```

### 11.2.2 Importing Data

```{r}
exam <- read_csv("../../data/Exam_data.csv")
```

## 11.3 Visualising Uncertainty of Point Estimates: ggplot2

### 11.3.1 Deriving Summary Statistics

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n = n(),
    mean = mean(MATHS),
    sd = sd(MATHS),
    .groups = "drop"
  ) %>%
  mutate(se = sd / sqrt(n - 1))

knitr::kable(head(my_sum), format = "html")
```

### 11.3.2 Plotting Standard Error Bars

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x = RACE,
        ymin = mean - se,
        ymax = mean + se),
    width = 0.2,
    colour = "black",
    alpha = 0.9,
    linewidth = 0.5) +
  geom_point(aes(x = RACE,
                 y = mean),
             stat = "identity",
             color = "red",
             size = 1.5,
             alpha = 1) +
  ggtitle("Standard error of mean maths score by race")
```

### 11.3.3 Plotting 95% Confidence Interval

```{r}
ggplot(my_sum) +
  geom_errorbar(
    aes(x = reorder(RACE, -mean),
        ymin = mean - 1.96 * se,
        ymax = mean + 1.96 * se),
    width = 0.2,
    colour = "black",
    alpha = 0.9,
    linewidth = 0.5) +
  geom_point(aes(x = RACE,
                 y = mean),
             stat = "identity",
             color = "red",
             size = 1.5,
             alpha = 1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

### 11.3.4 Interactive Error Bars

```{r}
shared_df <- SharedData$new(my_sum)

bscols(
  widths = c(4, 8),
  ggplotly(
    ggplot(shared_df) +
      geom_errorbar(
        aes(x = reorder(RACE, -mean),
            ymin = mean - 2.58 * se,
            ymax = mean + 2.58 * se),
        width = 0.2,
        colour = "black",
        alpha = 0.9,
        size = 0.5) +
      geom_point(
        aes(x = RACE,
            y = mean,
            text = paste("Race:", RACE,
                         "<br>N:", n,
                         "<br>Avg. Scores:", round(mean, digits = 2),
                         "<br>99% CI:[",
                         round(mean - 2.58 * se, digits = 2), ",",
                         round(mean + 2.58 * se, digits = 2), "]")),
        stat = "identity",
        color = "red",
        size = 1.5,
        alpha = 1) +
      xlab("Race") +
      ylab("Average Scores") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45,
                                       vjust = 0.5,
                                       hjust = 1)) +
      ggtitle("99% Confidence interval of average maths scores by race"),
    tooltip = "text"
  ),
  DT::datatable(shared_df,
                rownames = FALSE,
                class = "compact",
                width = "100%",
                options = list(pageLength = 10,
                               scrollX = TRUE),
                colnames = c("No. of pupils",
                             "Avg Scores",
                             "Std Dev",
                             "Std Error")) %>%
    formatRound(columns = c("mean", "sd", "se"),
                digits = 2)
)
```

## 11.4 Visualising Uncertainty: ggdist Methods

### 11.4.1 `stat_pointinterval()`

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval() +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Mean Point + Multiple-interval plot")
```

### 11.4.2 Customising `stat_pointinterval()`

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(.width = 0.95,
                     .point = median,
                     .interval = qi) +
  labs(
    title = "Visualising confidence intervals of median math score",
    subtitle = "Median Point + qi interval")
```

### 11.4.3 `stat_gradientinterval()`

```{r}
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(
    fill = "skyblue",
    show.legend = TRUE
  ) +
  labs(
    title = "Visualising confidence intervals of mean math score",
    subtitle = "Gradient + interval plot")
```

## 11.5 Hypothetical Outcome Plots (HOPs)

### 11.5.1 Installing ungeviz

```{r}
#| eval: false
devtools::install_github("wilkelab/ungeviz")
```

### 11.5.2 HOPs Plot

```{r}
if (requireNamespace("ungeviz", quietly = TRUE)) {
  library(ungeviz)

  ggplot(data = exam,
         aes(x = factor(RACE),
             y = MATHS)) +
    geom_point(position = position_jitter(height = 0.3,
                                          width = 0.05),
               size = 0.4,
               color = "#0072B2",
               alpha = 1 / 2) +
    geom_hpline(data = sampler(25,
                               group = RACE),
                height = 0.6,
                color = "#D55E00") +
    theme_bw() +
    transition_states(.draw, 1, 3)
} else {
  cat("`ungeviz` is not installed. Run `devtools::install_github(\"wilkelab/ungeviz\")` to render the HOPs example.")
}
```
