---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics"
author: "Jin Zhang"
date: "8 May 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: false
---

## 1.1 Learning Outcome

In this exercise, we learn the basic principles and essential components of **ggplot2** to create elegant and functional statistical graphics based on the Layered Grammar of Graphics.

## 1.2 Getting Started

### 1.2.1 Installing and Loading Packages

```{r}
pacman::p_load(tidyverse)
```

### 1.2.2 Importing Data

```{r}
exam_data <- read_csv("../../data/Exam_data.csv")
```

## 1.3 R Graphics vs ggplot2

```{r}
hist(exam_data$MATHS)
```

```{r}
ggplot(data = exam_data, aes(x = MATHS)) +
  geom_histogram(bins = 10,
                 boundary = 100,
                 color = "black",
                 fill = "grey") +
  ggtitle("Distribution of Maths scores")
```

## 1.4 Essential Elements: data

```{r}
ggplot(data = exam_data)
```

## 1.5 Essential Elements: Aesthetic Mappings

```{r}
ggplot(data = exam_data,
       aes(x = MATHS))
```

## 1.6 Essential Elements: geom

### 1.6.1 geom_bar

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

### 1.6.2 geom_dotplot

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

### 1.6.3 geom_histogram

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

### 1.6.4 Modify geom() - change fill colour by gender

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           fill = GENDER)) +
  geom_histogram(bins = 20,
                 color = "grey30")
```

### 1.6.5 geom_density

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           colour = GENDER)) +
  geom_density()
```

### 1.6.6 geom_boxplot

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot()
```

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot(notch = TRUE)
```

### 1.6.7 geom_violin

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_violin()
```

### 1.6.8 geom_point - scatter plot

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point()
```

### 1.6.9 Combining geom objects

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  geom_point(position = "jitter",
             size = 0.5)
```

## 1.7 Essential Elements: stat

### 1.7.1 stat_summary - add mean

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",
               fun = "mean",
               colour = "red",
               size = 4)
```

### 1.7.2 Using geom() method with stat

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  geom_point(stat = "summary",
             fun = "mean",
             colour = "red",
             size = 4)
```

### 1.7.3 Best fit curve with geom_smooth

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5)
```

## 1.8 Essential Elements: Facets

### 1.8.1 facet_wrap

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ CLASS)
```

### 1.8.2 facet_grid

```{r}
ggplot(data = exam_data,
       aes(y = MATHS,
           x = CLASS)) +
  geom_boxplot() +
  facet_grid(GENDER ~ .)
```

## 1.9 Essential Elements: Coordinates

### 1.9.1 Flip axes with coord_flip

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip()
```

### 1.9.2 Change axis range with coord_cartesian

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))
```

## 1.10 Essential Elements: Themes

### 1.10.1 theme_gray (default)

```{r}
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```
