---
title: "Hands-on Exercise 3: Interactive and Animated Data Visualisation with R"
author: "Jin Zhang"
date: "April 30, 2026"
date-modified: "last-modified"
execute:
  echo: true
  warning: false
  message: false
  freeze: false
---

## Loading R Packages

```{r}
pacman::p_load(ggiraph, plotly, patchwork, DT, tidyverse,
               readxl, gganimate, gifski, gapminder)
```

---

# Part 1: Interactive Data Visualisation with ggiraph

## Importing Data

```{r}
exam_data <- read_csv("../../data/Exam_data.csv")
```

## Interactive Dot Plot with Tooltip

`ggiraph` converts ggplot2 graphics into interactive SVG objects. We replace standard geoms with their `_interactive` counterparts and provide a `tooltip` aesthetic.

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p,
       width_svg = 6,
       height_svg = 6 * 0.618)
```

## Hover Highlighting with data_id

`data_id` groups elements so hovering over one point highlights all points sharing the same value — useful for seeing class groupings at a glance.

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p,
       width_svg = 6,
       height_svg = 6 * 0.618)
```

## Combining Tooltip and Hover Effect

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = CLASS,
        data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL, breaks = NULL)

girafe(ggobj = p,
       width_svg = 6,
       height_svg = 6 * 0.618)
```

## Coordinated Multiple Views

Linking two plots so hovering over a point in one plot simultaneously highlights the same student in the other.

```{r}
p1 <- ggplot(data = exam_data,
             aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL)

p2 <- ggplot(data = exam_data,
             aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL, breaks = NULL)

girafe(code = print(p1 + p2),
       width_svg = 6,
       height_svg = 3,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
       ))
```

---

# Part 2: Interactive Data Visualisation with plotly

## Basic Interactive Scatter Plot

`plot_ly()` creates interactive web-based visualizations. Hovering reveals values; clicking the legend toggles groups.

```{r}
plot_ly(data = exam_data,
        x = ~MATHS,
        y = ~ENGLISH,
        color = ~RACE)
```

## ggplotly()

`ggplotly()` wraps any existing ggplot2 object to add interactivity with minimal extra code.

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS,
                y = ENGLISH,
                color = RACE)) +
  geom_point(size = 1) +
  theme(legend.position = "bottom")

ggplotly(p)
```

## Coordinated Multiple Views with subplot()

Using `highlight_key()` and `subplot()` enables linked brushing — selecting points in one panel highlights the same records in the other.

```{r}
d <- highlight_key(exam_data)

p1 <- ggplot(data = d,
             aes(x = MATHS,
                 y = ENGLISH)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

p2 <- ggplot(data = d,
             aes(x = MATHS,
                 y = SCIENCE)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

subplot(ggplotly(p1),
        ggplotly(p2))
```

## Interactive Data Table with DT

```{r}
DT::datatable(exam_data, class = "compact")
```

---

# Part 3: Animated Data Visualisation with gganimate

## Importing Population Data

```{r}
col <- c("Country", "Continent")

globalPop <- read_xls("../../data/GlobalPopulation.xls",
                      sheet = "Data") %>%
  mutate(across(all_of(col), as.factor)) %>%
  mutate(Year = as.integer(Year))
```

## Static Bubble Plot

Before adding animation, we build the base ggplot2 bubble chart mapping Old (% aged) to x, Young (% young) to y, and Population to point size.

```{r}
ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_size(range = c(2, 12)) +
  labs(title = "Global Population: % Aged vs % Young",
       x = "% Aged",
       y = "% Young")
```

## Animated Bubble Plot with gganimate

`transition_time()` animates the chart across the `Year` variable. The title `{frame_time}` updates automatically each frame.

```{r}
ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_size(range = c(2, 12)) +
  labs(title = "Year: {frame_time}",
       x = "% Aged",
       y = "% Young") +
  transition_time(Year) +
  ease_aes("linear")
```

## Animated Bubble Plot with plotly

Using the `frame` aesthetic in `plot_ly()` adds a play button and year slider for interactive, user-controlled animation.

```{r}
globalPop %>%
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          frame = ~Year,
          text = ~Country,
          hoverinfo = "text",
          type = "scatter",
          mode = "markers") %>%
  layout(showlegend = TRUE,
         xaxis = list(title = "% Aged",
                      zeroline = FALSE),
         yaxis = list(title = "% Young",
                      zeroline = FALSE))
```
