---
title: "Hands-on Exercise 3: Interactive and Animated Data Visualisation"
author: "Jin Zhang"
date: "8 May 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: false
---

## 3.1 Learning Outcome

In this exercise we learn to create:

- Interactive visualisations using **ggiraph** and **plotly**
- Animated visualisations using **gganimate**
- Linked views using **crosstalk** and **DT**

## Part 1: Interactive Data Visualisation

## 3.2 Getting Started

### 3.2.1 Installing and Loading Packages

```{r}
pacman::p_load(ggiraph, plotly,
               patchwork, DT, tidyverse)
```

### 3.2.2 Importing Data

```{r}
exam_data <- read_csv("../../data/Exam_data.csv")
```

## 3.3 Interactive Visualisation - ggiraph methods

### 3.3.1 Tooltip effect

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6 * 0.618
)
```

### 3.3.2 Displaying multiple information on tooltip

```{r}
exam_data$tooltip <- c(paste0(
  "Name = ", exam_data$ID,
  "\n Class = ", exam_data$CLASS))
```

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = exam_data$tooltip),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 8,
  height_svg = 8 * 0.618
)
```

### 3.3.3 Customising tooltip style

```{r}
tooltip_css <- "background-color:white;
font-style:bold; color:black;"
```

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6 * 0.618,
  options = list(
    opts_tooltip(css = tooltip_css))
)
```

### 3.3.4 Displaying statistics on tooltip

```{r}
tooltip <- function(y, ymax, accuracy = .01) {
  mean <- scales::number(y, accuracy = accuracy)
  sem  <- scales::number(ymax - y, accuracy = accuracy)
  paste("Mean maths scores:", mean, "+/-", sem)
}
```

```{r}
gg_point <- ggplot(data = exam_data,
                   aes(x = RACE)) +
  stat_summary(aes(y = MATHS,
                   tooltip = after_stat(
                     tooltip(y, ymax))),
               fun.data = "mean_se",
               geom = GeomInteractiveCol,
               fill = "light blue") +
  stat_summary(aes(y = MATHS),
               fun.data = mean_se,
               geom = "errorbar",
               width = 0.2,
               size = 0.2)

girafe(ggobj = gg_point,
       width_svg = 8,
       height_svg = 8 * 0.618)
```

### 3.3.5 Hover effect with data_id

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6 * 0.618
)
```

### 3.3.6 Styling hover effect

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6 * 0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

### 3.3.7 Combining tooltip and hover effect

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(tooltip = CLASS,
        data_id = CLASS),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(
  ggobj = p,
  width_svg = 6,
  height_svg = 6 * 0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),
    opts_hover_inv(css = "opacity:0.2;")
  )
)
```

### 3.3.8 Coordinated multiple views with ggiraph

```{r}
p1 <- ggplot(data = exam_data,
             aes(x = MATHS)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)

p2 <- ggplot(data = exam_data,
             aes(x = ENGLISH)) +
  geom_dotplot_interactive(
    aes(data_id = ID),
    stackgroups = TRUE,
    binwidth = 1,
    method = "histodot") +
  coord_cartesian(xlim = c(0, 100)) +
  scale_y_continuous(NULL,
                     breaks = NULL)

girafe(code = print(p1 + p2),
       width_svg = 10,
       height_svg = 5,
       options = list(
         opts_hover(css = "fill: #202020;"),
         opts_hover_inv(css = "opacity:0.2;")
       ))
```

## 3.4 Interactive Visualisation - plotly methods

### 3.4.1 plot_ly() - basic scatter plot

```{r}
plot_ly(data = exam_data,
        x = ~MATHS,
        y = ~ENGLISH)
```

### 3.4.2 plot_ly() - with visual variable (colour by Race)

```{r}
plot_ly(data = exam_data,
        x = ~MATHS,
        y = ~ENGLISH,
        color = ~RACE)
```

### 3.4.3 ggplotly() method

```{r}
p <- ggplot(data = exam_data,
            aes(x = MATHS,
                y = ENGLISH)) +
  geom_point(dotsize = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

ggplotly(p)
```

### 3.4.4 Coordinated multiple views with plotly

```{r}
d <- highlight_key(exam_data)

p1 <- ggplot(data = d,
             aes(x = MATHS,
                 y = ENGLISH)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

p2 <- ggplot(data = d,
             aes(x = MATHS,
                 y = SCIENCE)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

subplot(ggplotly(p1),
        ggplotly(p2))
```

## 3.5 Interactive Data Table - DT package

```{r}
DT::datatable(exam_data,
              class = "compact")
```

## 3.6 Linked Brushing - crosstalk method

```{r}
d <- highlight_key(exam_data)

p <- ggplot(d,
            aes(ENGLISH,
                MATHS)) +
  geom_point(size = 1) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100))

gg <- highlight(ggplotly(p),
                "plotly_selected")

crosstalk::bscols(gg,
                  DT::datatable(d),
                  widths = 5)
```

## Part 2: Animated Data Visualisation

## 3.7 Getting Started for Animation

### 3.7.1 Installing and Loading Packages

```{r}
pacman::p_load(readxl, gganimate,
               gifski, gapminder,
               plotly, tidyverse)
```

### 3.7.2 Importing Population Data

```{r}
col <- c("Country", "Continent")

globalPop <- read_xls("../../data/GlobalPopulation.xls",
                      sheet = "Data") %>%
  mutate(across(all_of(col), as.factor)) %>%
  mutate(Year = as.integer(Year))
```

## 3.8 Animated Visualisation Methods

### 3.8.1 Static bubble plot

```{r}
ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = "Global Population: % Aged vs % Young",
       x = "% Aged",
       y = "% Young")
```

### 3.8.2 Animated bubble plot with gganimate

```{r}
ggplot(globalPop,
       aes(x = Old,
           y = Young,
           size = Population,
           colour = Country)) +
  geom_point(alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = "Year: {frame_time}",
       x = "% Aged",
       y = "% Young") +
  transition_time(Year) +
  ease_aes("linear")
```

### 3.8.3 Animated bubble plot with ggplotly

```{r}
p <- ggplot(globalPop,
            aes(x = Old,
                y = Young,
                colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7,
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = "% Aged",
       y = "% Young")

ggplotly(p)
```

### 3.8.4 Animated bubble plot with plotly

```{r}
globalPop %>%
  plot_ly(x = ~Old,
          y = ~Young,
          size = ~Population,
          color = ~Continent,
          frame = ~Year,
          text = ~Country,
          hoverinfo = "text",
          type = "scatter",
          mode = "markers") %>%
  layout(showlegend = TRUE,
         xaxis = list(title = "% Aged"),
         yaxis = list(title = "% Young"))
```
