---
title: "Hands-on Exercise 2: Beyond ggplot2 Fundamentals"
author: "Jin Zhang"
date: "8 May 2026"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  message: false
  freeze: false
---

## 2.1 Overview

In this exercise we explore ggplot2 extensions for more elegant and effective statistical graphics:

- **ggrepel** for non-overlapping text annotations
- **ggthemes** and **hrbrthemes** for professional themes
- **patchwork** for combining multiple plots

## 2.2 Getting Started

### 2.2.1 Installing and Loading Packages

```{r}
pacman::p_load(ggrepel, patchwork,
               ggthemes, hrbrthemes,
               tidyverse)
```

### 2.2.2 Importing Data

```{r}
exam_data <- read_csv("../../data/Exam_data.csv")
```

## 2.3 Beyond ggplot2 Annotation: ggrepel

### 2.3.1 Problem - overlapping labels with geom_label

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  geom_label(aes(label = ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### 2.3.2 Solution - geom_label_repel

```{r}
ggplot(data = exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold") +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## 2.4 Beyond ggplot2 Themes

### 2.4.1 Default theme_gray

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")
```

### 2.4.2 Working with ggthemes - The Economist theme

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

### 2.4.3 Working with hrbrthemes

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

### 2.4.4 hrbrthemes with axis and grid customisation

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## 2.5 Beyond Single Graph: patchwork

### 2.5.1 Create individual plots

```{r}
p1 <- ggplot(data = exam_data,
             aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(data = exam_data,
             aes(x = ENGLISH)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  coord_cartesian(xlim = c(0, 100)) +
  ggtitle("Distribution of English scores")

p3 <- ggplot(data = exam_data,
             aes(x = MATHS,
                 y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              linewidth = 0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) +
  ggtitle("English vs Maths scores")
```

### 2.5.2 Combining two plots side by side

```{r}
p1 + p2
```

### 2.5.3 Combining three plots

```{r}
(p1 / p2) | p3
```

### 2.5.4 Creating composite figure with tag

```{r}
((p1 / p2) | p3) +
  plot_annotation(tag_levels = "I")
```

### 2.5.5 Creating figure with inset

```{r}
p3 + inset_element(p2,
                   left = 0.02,
                   bottom = 0.7,
                   right = 0.5,
                   top = 1)
```

### 2.5.6 Composite figure with patchwork and ggtheme

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```
